Model Finding for (Co)recursive Functions in SMT

### 2015-04-30 tex/main.tex
Revisions, refutation-soundness argument.

\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\cvciii}{\textsc{cvc}{\small 3}\xspace}
\newcommand{\ziii}{\textsc{z}{\small 3}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\cc}[1]{#1^*}
\newcommand{\terms}{\mathbf{T}}
%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\M}{\mathcal{M}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
\newcommand{\conv}{\mathcal{W}}

@@

In this paper, we make the following contributions:
\begin{itemize}
\item[-] We define a satisfiability-preserving translation method for formulas containing recursive function definitions
that produces formulas for which existing model finding techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
\item[-] We provide experimental evidence that this translation improves the effectiveness of SMT solvers \cvc and \ziii for finding counterexamples to verification conditions.
\item[-] We detail front-end support for recursive function definitions and this translation in the SMT solver \cvc.
\end{itemize}

@@

In this paper, we associate quantified formulas with the uninterpreted functions whose semantics they define.
We will write $\forallf{f} \vec x. \varphi$ to denote a quantified formula that is defines the semantics of uninterpreted function $f$.
For consistency, we consider only quantified formulas of this form, 
and allow $f$ to be a distinguished constant $\fnull$, 
denoting that the quantified formula does not define the semantics of an uninterpreted function.
When $f$ is not $\fnull$, we will refer to $\forallf{f} \vec x. \varphi$ as a \emph{function definition} (for $f$).
We will write $\forall \vec x. \varphi$ as shorthand for $\forallf{\fnull} \vec x. \varphi$,
and $\exists \vec x. \varphi$ as shorthand for $\neg \forall \vec x. \neg \varphi$.

@@

Given a signature $\Sigma$, 
we will commonly parition its function symbols $\sfuns{\Sigma}$ into $\sfundefs{\Sigma} \mathrel{\uplus} \sfunndefs{\Sigma}$,
where $\sfundefs{\Sigma}$ is the set of \emph{defined} function symbols,
and $\sfunndefs{\Sigma}$ are the others,
and consider $\Sigma$-formulas that are well-defined with respect to functions $\sfundefs{\Sigma}$.

@@

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ : 
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi, D ) = \conv( \varphi, \ptrue )$;
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] $\mathsf{return}$ $\psi$
  \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv_0( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$;
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv_0( t_1, p )$;
      \item[] $\mathsf{if}$ $f \equiv \fnull$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        %\item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \terms( t ), f \in \sfundefs{\Sigma} \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to a set of well-specified function definitions $\sfundefs{\Sigma}$.
In the subprocedure $\conv_0$, the procedure $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an (interpreted) predicate.
}
\label{fig:encoding}
\end{figure}

In more detail, 
say our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
We extend the signature of $\Sigma$ to a signature $\extendsig{\Sigma}$ containing the following additional definitions.
For each $f \in \sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$, 
the signature $\extendsig{\Sigma}$ includes:
\begin{enumerate}
\item[-] an uninterpreted type $\fargsort{f}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ of type $(\fargsort{f} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{f} \rightarrow \tau_n)$.
\end{enumerate}
%The interpretation of uninterpreted sort $\fargsort{f}$ will denote the elements (tuples) on which the function $f$ is applied.
%The role of the uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ will be discussed more in the following.

@@

In the case that $t$ is a quantified formula $\forallf{f} \vec x. t_1$, it recursively calls $\conv$ on its body with the same polarity.
In the case that $t$ is not a function definition, 
the resulting quantified formula is reconstructed with the term $r_1$ returned by the recursive call,
and a quantified prefix is appended to the formula $D_1$ returned by the recursive call.
In the case that $t$ is a function definition,
we instead construct a quantified formula over a single variable $y$ of type $\fargsort{f}$,
and replace all occurrences of $\vec x$ in $r_1$ with $\vecfarg{f}( y )$.
In this case, since function definitions are a top-level conjunct, by case analysis on the return values of $\conv$, we know that $D_1$ is empty.

@@

\begin{example}
\label{ex:translation}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$ and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{eqnarray} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{eqnarray} 
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the type $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of type $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
The universal quantification in the resulting formula is over an uninterpreted type $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
We observe that both \cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}),
and both produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.
$\square$
\end{example}

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\M$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{eqnarray}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{eqnarray}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\M$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\M$ interprets e.g. $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are well-specified, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp. non-existence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For instance, the model $\M$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$, 
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with a model of $\conv( \varphi )$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is in fact a non-issue, as $s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.

@@

\begin{theorem}
If $\varphi$ be a $\Sigma$-formula that is well-defined with respect to functions $\sfundefs{\Sigma}$,
and the set of function definitions corresponding to functions in $\sfundefs{\Sigma}$ is well-specified,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}

@@

When reading this input, 
\cvc adds the (annotated) quantified formula $\forallf{s} x. s( x ) \teq \lite( x \leq 0, 0, s( x-1 ) )$ to its list of assertions,
which after rewriting becomes $\forallf{s} x. \lite( x \leq 0, s( x ) \teq 0, s( x ) \teq s( x-1 ) )$.
If \cvc is in finite model finding mode for recursive functions is enabled (using the command line parameter ``--fmf-fun"),
then it will replace its list of known assertions based on the conversion $\conv$ before checking for satisfiability.
We stress that this option is model-sound only if the functions provided in the input are well-specified.

>>>>

\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\cvciii}{\textsc{cvc}{\small 3}\xspace}
\newcommand{\ziii}{\textsc{z}{\small 3}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\cc}[1]{#1^*}
\newcommand{\terms}{\mathbf{T}}
%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
\newcommand{\conv}{\mathcal{W}}

@@

In this paper, we make the following contributions:
\begin{itemize}
\item[-] We define a satisfiability-preserving translation method for a class of formulas containing recursive function definitions
that produces formulas for which existing model finding techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
\item[-] We provide experimental evidence that this translation improves the effectiveness of SMT solvers \cvc and \ziii for finding counterexamples to verification conditions.
\item[-] We detail front-end support for recursive function definitions and this translation in the SMT solver \cvc.
\end{itemize}

@@

%In this paper, we associate quantified formulas with uninterpreted functions (informally, those whose semantics they define).
In this paper, we consider \emph{annotated quantifed formulas} of the form $\forallf{f} \vec x. \varphi$, 
where $f$ is an uninterpreted function from $\sfuns{\Sigma}$, or the distinguished constant $\fnull$.
When $f$ is not $\fnull$, we will refer to $\forallf{f} \vec x. \varphi$ as a \emph{function definition} (for $f$).
We will write $\forall \vec x. \varphi$ as shorthand for $\forallf{\fnull} \vec x. \varphi$,
and $\exists \vec x. \varphi$ as shorthand for $\neg \forall \vec x. \neg \varphi$.

@@

Given a signature $\Sigma$, 
we will commonly parition its function symbols $\sfuns{\Sigma}$ into $\sfundefs{\Sigma} \mathrel{\uplus} \sfunndefs{\Sigma}$,
where $\sfundefs{\Sigma}$ is the set of \emph{defined} function symbols,
and consider $\Sigma$-formulas that are well-defined with respect to functions $\sfundefs{\Sigma}$.

@@

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ : 
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi, D ) = \conv( \varphi, \ptrue )$;
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] $\mathsf{return}$ $\psi$
  \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv_0( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$;
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv_0( t_1, p )$;
      \item[] $\mathsf{if}$ $f \in \sfundefs{\Sigma}$
      \begin{itemize}
        %\item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \terms( t ), f \in \sfundefs{\Sigma} \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to a set of well-specified function definitions $\sfundefs{\Sigma}$.
In the subprocedure $\conv_0$, the procedure $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an interpreted predicate.
}
\label{fig:encoding}
\end{figure}

In more detail, 
say our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
We extend the signature of $\Sigma$ to a signature $\extendsig{\Sigma}$ containing the following additional definitions.
For each $f \in \sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \times \tau_n \rightarrow \tau$, 
the signature $\extendsig{\Sigma}$ includes:
\begin{enumerate}
\item[-] an uninterpreted type $\fargsort{f}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ of type $(\fargsort{f} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{f} \rightarrow \tau_n)$.
\end{enumerate}
%The interpretation of uninterpreted sort $\fargsort{f}$ will denote the elements (tuples) on which the function $f$ is applied.
%The role of the uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ will be discussed more in the following.

@@

In the case that $t$ is a quantified formula $\forallf{f} \vec x. t_1$, it recursively calls $\conv$ on its body with the same polarity.
In the case that $t$ is a function definition of a function in $\sfundefs{\Sigma}$,
we instead construct a quantified formula over a single variable $y$ of type $\fargsort{f}$,
and replace all occurrences of $\vec x$ in $r_1$ with $\vecfarg{f}( y )$.
In this case, since function definitions are a top-level conjunct, by case analysis on the return values of $\conv$, we know that $D_1$ is empty.
In the case that $t$ is not a function definition, 
the resulting quantified formula is reconstructed with the term $r_1$ returned by the recursive call,
and a quantified prefix is appended to the formula $D_1$ returned by the recursive call.

Otherwise, $t$ is either an application of an uninterpreted predicate, or a term of a type other than $\Bool$.
In this case, the second component of our return value contains a constraint of the form $\exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t )$
for each subterm $f( \vec t )$ of $t$ such that $f$ is a function in $\sfundefs{\Sigma}$.
When such constraints are asserted positively, this indicates that the tuple of uninterpreted functions $\vecfarg{f}$ must include $\vec t$ in its range.

@@

\begin{example}
\label{ex:translation}
Consider the (combined) signature of linear arithmetic and uninterpreted functions $\Sigma = \Sigmalia \cup \Sigma_{\euf}$,
where $\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and 
$\sfuns{\Sigma_{\euf}} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{eqnarray} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{eqnarray} 
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the type $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of type $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
The universal quantification in the resulting formula is over an uninterpreted type $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
We observe that both \cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}),
and both produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.
$\square$
\end{example}

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{eqnarray}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{eqnarray}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\I$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ interprets e.g. $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are well-specified, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$ for any input $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ in the previous example is not relevant to the satisfiability of formula~(\ref{eq:ex-before}).
Thus, a satisfiable (resp. unsatisfiable) response from an SMT solver on input $\conv( \varphi )$ implies the existence (resp. non-existence) of a model for $\varphi$.
Moreover, information in models for $\conv( \varphi )$ contains relevant information regarding the models of $\varphi$.
For instance, the model $\I$ for formula~(\ref{eq:ex-after}) described above interprets $c$ as $14$, 
while there exists a model of formula~(\ref{eq:ex-before}) also interprets $c$ as $14$.
In fact, we claim that there exists a model of $\varphi$ that coincides with each model of $\conv( \varphi )$ on its interpretation of all functions in $\sfuns{\Sigma}$ 
other than those in $\sfundefs{\Sigma}$, i.e. $\{ s \}$ in the previous example.
From a practical point of view, this is in fact a non-issue, as $s$ is the very function that was explicitly defined by the user,
and hence already has an intended interpretation.

@@

%\begin{lemma}
%Let $\M$ be a model satisfying $\conv( \varphi )$, and
%%and let $\conv( \varphi )^\ast$ be the quantifier-free formula obtained by 
%let $M$ be a set of literals such that $\M \models M$ and $M \models_p \conv( \varphi )$.
%For each assignment $\sigma$ over $\vars{ M }$,
%each $f( \vec t ) \in \terms( M \sigma )$ where $f \in \sfundefs{\Sigma}$, 
%there exists a formula of the form $\exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t )$ in $M \sigma$.
%\end{lemma}

\begin{theorem}
If $\varphi$ be a $\Sigma$-formula that is well-defined with respect to functions $\sfundefs{\Sigma}$,
and the set of function definitions corresponding to functions in $\sfundefs{\Sigma}$ is well-specified,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ has a model $\I$, then $\conv( \varphi )$ has a model $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all sorts $\sigma \in \ssorts{\Sigma}$ as $\sigma^\I$,
all functions $f \in \sfuns{\Sigma}$ as $f^\I$,
and for each function $f$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{f}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\farg{f}^i$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t )$,
since by our construction of $\J$ there exists a $z$ such that $z^\J = \vec t^\J$.
Since $\I$ satisfies $\varphi$ and due to our construction of $\conv( \varphi )$, we conclude $\J$ is a model of $\conv( \varphi )$.

Second, we show that if $\conv( \varphi )$ has a model $\I$, then $\varphi$ has a model $\J$.

\end{proof}

@@

When reading this input, 
\cvc adds the (annotated) quantified formula $\forallf{s} x. s( x ) \teq \lite( x \leq 0, 0, s( x-1 ) )$ to its list of assertions,
which after rewriting becomes $\forallf{s} x. \lite( x \leq 0, s( x ) \teq 0, s( x ) \teq s( x-1 ) )$.
If \cvc is in finite model finding mode for recursive functions is enabled (using the command line parameter ``--fmf-fun"),
then it will replace its list of known assertions based on the conversion $\conv$ before checking for satisfiability.
Note that this option is model-sound only if the functions provided in the input are well-specified.
